---
title: "Reproducible Research: Peer Assessment 1"
author: "B. Bansenauer"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r, echo=FALSE, results='hide'}
unzip("activity.zip") # unzip files
df = read.csv("activity.csv")
file.remove("activity.csv") # tidy up by removing the csv
```
## What is mean total number of steps taken per day?
```{r echo=-1}
library(plyr)
df.stepsperday<-ddply(df, .(date), summarize, total= sum(steps))
mean(na.omit(df.stepsperday$total))
median(na.omit(df.stepsperday$total))
hist(df.stepsperday$total, 
     main="Total Steps per Day",
     xlab= "Number of Steps")
```

## What is the average daily activity pattern?
Plot the average number of steps taken across all days during each 5-minute interval.
```{r echo=2:3}
library(ggplot2)
df.intervals<-ddply(df, .(interval), summarize, mean=mean(na.omit(steps)))
ggplot(df.intervals, aes(interval, mean))+geom_line()+ylab("Average Number of Steps across All Days")
```

Investigate which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
df.intervalMax<-subset(df.intervals, mean==max(df.intervals$mean, na.rm = TRUE))
```
**Interval `r df.intervalMax$interval` contains the maximum average steps of `r df.intervalMax$mean`.**

## Inputing missing values
1. Calculate and report the total number of missing values in the dataset
(i.e. the total number of rows with NAs)
```{r}
count(is.na(df))
```

3. Create a new dataset that is equal to the original dataset but with the missing step data filled in with the 5-minute interval mean (rounded).
```{r}
df.intervalstack<-rep(df.intervals$mean,61)
df.na<-is.na(df$steps)
df.filled<-df
df.filled$steps[df.na]<-round(df.intervalstack[df.na])
```
4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. 
```{r}
df.stepsperday.filled<-ddply(df.filled, .(date), summarize, total= sum(steps))
hist(df.stepsperday.filled$total, 
     main="Total Steps per Day",
     xlab= "Number of Steps")
```

**The total number of steps per day using the filled dataset shows a slightly lower mean and median:**
```{r}
mean(df.stepsperday.filled$total) 
median(df.stepsperday.filled$total)
```


## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r echo=2:3, message=FALSE}
library(dplyr)
df2 <- mutate(df.filled, day = factor(case_when(weekdays(as.Date(date),abbreviate=TRUE) %in% c("Sat", "Sun") ~ "weekend",  TRUE ~ "weekday")))
summary(df2)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekdays or weekend days (y-axis). 
```{r echo=-(3:4)}
df2.weekday.intervals<-ddply(subset(df2, day=="weekday"), .(interval), summarize, mean=mean(steps))
df2.weekend.intervals<-ddply(subset(df2, day=="weekend"), .(interval), summarize, mean=mean(steps))
library(grid)
library(gtable)
p1<-ggplot(df2.weekday.intervals, aes(interval, mean))+geom_line()+labs(title = "Weekdays",  x="5-Minute Interval", y = "Average Number of Steps")
p2<-ggplot(df2.weekend.intervals, aes(interval, mean))+geom_line()+labs(title= "Weekend Days", x="5-Minute Interval", y = "Average Number of Steps")
g1 <- ggplotGrob(p1)
g2 <- ggplotGrob(p2)
g <- rbind(g2, g1, size = "first")
g$widths <- g1$widths
grid.newpage()
grid.draw(g)
```
